Learning how to act: making good decisions with machine learning

### 2017-08-24 thesis/thesis_map.tex
begins restructure to two intro chapters

\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}

@@

@@

@@

\chapter{Causal models}
\label{chap:causal_models}

@@

Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention. The power of defining an intervention in this way stems from the number of things that are invariant between the two processes. All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.

@@

\subsubsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

Their difference is also normal. Let $\tau = \cf{Y}{1} - \cf{Y}{0}$,

@@

\section{Granger causality}

Mentioned in the category of causal frameworks,

Granger causality is less of a framework for defining causality and on which a wide range of algorithms and methods can be constructed, but a method for infering causal structure from data. As such is discussion really belongs in section \ref{sec:causal-discovery}.

TODO move this to a paragraph at the end of the comparing and unifying section. \todo{}

@@

\chapter{Two key questions}
\label{chap:two_key_questions}

We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery. In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known. The goal is then to estimate the effect of an intervention or range of interventions in the system. Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}. Almost every time someone runs a regression model the key quantity of interest is a causal effect. Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.

Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data. This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another. Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}. An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.

\section{Causal effect estimation} 
\label{sec:causal effect estimation}

@@

\section{Causal Discovery}
\label{sec:causal-discovery}

%Cyclic models \citep{Hyttinen2012}

We now move to the much more general problem of learning a causal graph from observational data. In this setting we make much broader assumptions about the structure of the graph. For example, that it is acyclic or that we have no unmeasured confounding variables. We do not assume the existence or directions of any links between the variables. Amazingly, it is possible to infer some aspects of causal structure with such general assumptions. The set of conditional independence in a non-experimental data set indicates some causal structures are more likely than others. In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause. These clues are the key to causal discovery algorithms.

%There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.

Causal discovery is a much grander goal than causal effect estimation given a known causal network. Arguably, if achieved, it would equate to the automation of scientific discovery. We need simply supply our algorithm with a vast collection of variables (regardless of their relevance to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make. Unfortunately, causal discovery is very hard. Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows exponentially with the square of the number of variables .

% - some intuition here about the strength of the signal we get from causal discovery. Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}. Talk about the numerical as well as the statistical issues here.

In the next sections we briefly survey the key approaches to causal discovery. We roughly divide the methods into those based on those that exploit the connection between the conditional independencies in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.

%This is not the only way we could have grouped them. There is also a separation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best. There are also hybrid methods \ref{}

\subsection{Conditional independence based methods}
One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution. For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network. If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$. However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network. This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$. This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$. Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \citep{pearl1988probabilistic} for $P$.

Faithfulness is an assumption. It does not always hold and we cannot verify it from the observational data we wish to use for causal inference. However, most distributions generated by a causal Bayesian network will be faithful to that network. For faithfulness to be violated, different causal effects must exactly balance one-another out. For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure). If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption. However, this independence is not stable. It would disappear under a small perturbation to any of the parameters. In discrete systems, violations correspond to the solutions to polynomial equations over values in the CPD tables and thus are a space of measure zero with respect to all possible distributions associated with the graph \citep{Koller2009}.

Given the faithfulness assumption, our causal discovery problem reduces to finding the set of Bayesian networks that have exactly the dependency structure as we observe in $P$. This set can also be referred to as the Markov equivalence class compatible with $P$.

\subsubsection{Without hidden common causes}
The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of Bayesian networks we must consider. This assumption is referred to as causal sufficiency \citep{Sprites2000}.

We will begin with a brute force algorithm (described as the SGS algorithm in \citet{Sprites2000} and IC algorithm in \citet{Pearl2000}). While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.

\begin{table}[H]
 \begin{tabularx}{\textwidth}{X}
 \hline
\rule{0pt}{2.5ex} 
 \textbf{The SGS (or IC) Algorithm}\\
 \hline
 \rule{0pt}{2.5ex}
\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
 \begin{enumerate}[itemsep=8pt]
  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an un-directed link to form a complete graph.
  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent. If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
  \label{alg:SGSexponential}

  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
  \label{alg:SGScolliders}  
  \item Recursively try to orient any edges that remain un-directed to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
  \label{alg:SGSfinal}
\end{enumerate}\\
 \hline
\end{tabularx}
\end{table}

The SGS algorithm utilises the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation. Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model. Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables. Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables. If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.

The PC algorithm \citet{Sprites2000} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$. It also checks for low order conditional independence relations before higher order ones. This allows it to exploit any sparsity in the true network, leading to much better average case performance  \citep{Sprites2000} (although the worst case, where the true network is complete, is still exponential). With finite data, the order in which the links are considered can change the output (unlike for SGS). The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.

\begin{table}[H]
 \begin{tabularx}{\textwidth}{X}
 \hline
\rule{0pt}{2.5ex} 
 \textbf{The PC Algorithm}\\
 \hline
 \rule{0pt}{2.5ex}
\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
 \begin{enumerate}[itemsep=8pt]
  \item As for SGS
  \item \textbf{for} each link $a-b$:
  \begin{itemize}[label={}]
   \item $n = 0$
   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
   \item 
    	\begin{itemize}[label={}]
    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
    	\item \begin{itemize}[label={}]
    			delete the link
    		  \end{itemize}

    	\item $n = n+1$
    	\end{itemize}
  \end{itemize}   
  \item as for SGS
  \item as for SGS
\end{enumerate}\\
 \hline
\end{tabularx}
\end{table}

The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution. Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects from the truncated product formula \ref{eq:truncatedproduct}. We can then bound the true causal effect by combining the results for the all the networks. This procedure is the IDA algorithm \citep{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \citep{Maathuis2010}. An implementation is available in the R package \citep{Kalisch2012}

\subsubsection{With hidden variables}
There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables. With an unknown number of hidden variables there are infinity many possible structures to search over. In addition, the space of causal networks is not closed under marginalisation. If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any Bayesian network (see figure \ref{fig:DAGSnotclosed}). The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables).

\begin{figure}
\centering
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={observed},lt/.style={latent}]
 %nodes
\node[lt](1){U}; 
\node[main node, below left = of 1](2){B};
\node[main node, below right = of 1](3){C};
\node[main node, above left= of 2](4){A};
\node[main node, above right = of 3](5){D};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    	edge node {} (3)
    (4) edge node {} (2)
    (5) edge node {} (3);

\end{tikzpicture}
\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
\label{fig:DAGSnotclosed}
\end{figure}

\begin{theorem}
\citet{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
\end{theorem}

Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables. Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable. These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalisation and conditioning.

For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}. This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites2000}. The logic behind the algorithm is very similar. Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations. The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \citep{Spirtes1995}. Assuming there are no selection variables, the PAG can contain four types of link:

\begin{enumerate}
\item $X \rightarrow Y$, meaning $X$ causes $Y$
\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
\end{enumerate}

The circles indicate where it is ambiguous if there should be an arrowhead (IE where there is one in some MAGs and not in others in the equivalence class). Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).

\begin{figure}
\centering
 \begin{subfigure}[t]{0.4\textwidth}
 \centering
 \caption{True DAG}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed},lt/.style={latent}]
 %nodes
\node[lt](1){U}; 
\node[main node, below left = of 1](2){B};
\node[main node, below right = of 1](3){C};
\node[main node, above left= of 2](4){A};
\node[main node, above right = of 3](5){D};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    	edge node {} (3)
    (4) edge node {} (2)
    (5) edge node {} (3);

\end{tikzpicture}
 \end{subfigure}
 \begin{subfigure}[t]{0.4\textwidth}
  \centering
  \caption{Partial ancestral graph}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed}]
 %nodes

\node[main node](1){A};
\node[main node, below right=of 1](2){B};
\node[main node, right=of 2](3){C};
\node[main node, above right=of 3](4){D};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge[pil] node {} (2)
    (2) edge node {} (3)
    (3) edge node {} (2)
    (4) edge [pil] node {} (3);    	
\end{tikzpicture}
\end{subfigure}
\par\bigskip 
 \begin{subfigure}[t]{0.4\textwidth}
 \centering
 \caption{True DAG}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm, thick,main node/.style={observed},lt/.style={latent}]
 %nodes

\node[main node](1){C};
\node[main node, above left=of 1](2){A};
\node[main node, above right=of 1](3){B};
\node[main node, below=of 1](4){D};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (4)
    (2) edge node {} (1)
    (3) edge node {} (1);    	
\end{tikzpicture}
\end{subfigure}
 \begin{subfigure}[t]{0.4\textwidth}
  \centering
  \caption{Partial ancestral graph}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={observed}]
 %nodes

\node[main node](1){C};
\node[main node, above left=of 1](2){A};
\node[main node, above right=of 1](3){B};
\node[main node, below=of 1](4){D};

 \path[every node/.style={font=\sffamily\small}]
 	(1) edge node {} (4)
    (2) edge [pil] node {} (1)
    (3) edge [pil] node {} (1);    	
\end{tikzpicture}
\end{subfigure}
\caption{FCI examples: true graph and FCI output}
\label{fig:FCIExamples}
\end{figure}

The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \citep{Zhang2008}. More recently \citet{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster. \citet{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.

Latent variables can create constraints on the marginal distribution over the observed variables that cannot be expressed in term of conditional independencies. These generalised constraints can be expressed and leveraged within nested Markov models \citep{Richardson2012,Shipster2014}

%\renewcommand{\arraystretch}{1.5}
%\begin{tabular}{| c | c | p{4cm} | c | c | c |}
%\hline
%  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
%  \hline
%  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
%  \hline
%  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
%  \hline
%  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
%  \hline
%  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
%  \hline
%  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
%  \hline
%  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
% \hline
%\end{tabular}

All the algorithms discussed in this section rely on being able to perform conditional independence tests. This is non-trivial with high dimensional data. If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero \citep{Koller2009}. Where the functions are non-linear one can apply kernelised independence tests \citep{Gretton2008,Zhang2012}

\subsection{Discovery with functional models}
The algorithms we have considered so far return a Markov equivalence class. They cannot distinguish between two models that result in the same set of conditional independence relations. Consider the very simple case where  there are only two variables and the possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$. These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$. No algorithm relying purely on conditional independence relations can separate these two cases.

Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$. What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from? Recall the functional definition of causality (section \ref{sec:SEM}). There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \citep{Hoyer2009}, post-non-linear additive noise \citep{Zhang2008a} or linear models with non-Gaussian noise \citep{Hoyer2012}.

The causal direction can also be identified via a connection between casual discovery and semi-supervised learning \citep{Janzing2012}. Suppose we are trying to learn $\P{Y|X}$. The goal of semi-supervised learning is to improve our estimate of $\P{Y|X}$ by leveraging additional data sampled from $\P{X}$. However, if the true causal model is $X \rightarrow Y$ then their is some function mapping values of $X$ to $Y$ which should be invariant to any changes in the input distribution $\P{X}$. Therefor $\P{X}$ should be independent of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning. However if the true causal model is $Y \rightarrow X$ then variations in the $\P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help. This assumption of independence of mechanism and input can also allow the identification of the causal direction in SEMs even where the functions are deterministic and invertable \citep{Daniusis2010}. \citet{Janzing2012a} leverage an information geometric viewpoint of the independence of mechanism and input to infer the causal direction between a pair of associated variables.

Instead of positing a functional restriction on the relationship between variables and then developing theory to exploit that assumption, \citep{LopezPaz2014} propose learning what the causal relationship looks like from data. They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$. Their algorithm requires a data set in which each row is itself a data set consisting of pairs of variables $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$. They use a kernel mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-data set and train an algorithm to learn the direction of causality. Unfortunately we do not have a large collection of data sets where the causal direction is known to train such a model. \citet{LopezPaz2014} instead use a simulated data set so their model will necessarily be based on the assumptions they make when generating the data. Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.

\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting. They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable. They build on this to develop an algorithm that allows the construction of causal graphs based on the additive noise assumption.

\chapter{The interventionist viewpoint}

@@

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{y_1,...,y_k}$ be a joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate}
At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$. We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$. 
\end{definition}

The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled. To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step. This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature. There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice. In addition, researches often refer to the expected regret as \quotes{the regret}.}.
\eqn{
\label{eqn:raw_regret}
\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
}

@@

\vspace{0.5cm}
\begin{definition}[Pseudo-Regret]
\eqn{
\label{eqn:pseudo_regret}
\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}} \\
&= n\mu_{i^*} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
}
\end{definition}

@@

@@

It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}. However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem. \citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information. This is an example of the case represented in figure \ref{fig:context-graph-confounded}.

>>>>

\newcommand{\V}[1]{\mathbb V\left[#1\right]}
\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}

@@

\item What are the nessessary and sufficient conditioins for decision making? How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function). Ie when is the mean sufficient? What about the mean and the variance? - when can we take one set of results and compute something about a different loss function?

@@

-------------------

In chapter \ref{chap:interventionalist_view} I describe the interventionalist viewpoint on learning the outcome of actions.

I review the key problems and results within the bandit liturature

highlight some interesting connections between existing strands of work on bandits and causal inference.

-------------------

@@

\chapter{Causal inference from observational data}
\label{chap:causality}

\section{Causal models}
\label{sec:causal_models}

@@

\subsection{Can't everything be encoded in a Bayesian model?}
Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention. The power of defining an intervention in this way stems from the number of things that are invariant between the two processes. All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.

@@

\subsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

The distribution of \emph{individual treatment effects}, $\tau = \cf{Y}{1} - \cf{Y}{0}$ is also normal,

@@

@@

\section{Causal effect estimation} 
\label{sec:causal effect estimation}
We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery. In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known. The goal is then to estimate the effect of an intervention or range of interventions in the system. Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}. Almost every time someone runs a regression model the key quantity of interest is a causal effect. Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.

Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data. This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another. Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}. An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour. \todo{note that I focus on problems where causal structure known, reference rewiew on structure learning}

@@

\chapter{Learning from interventions}
\label{chap:interventionalist_view}

@@

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
\end{enumerate}

At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$. We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$. Note we have used counterfactual notation (see ) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have recieved if it selected action $i$ at timestep $t$.  
\end{definition}

The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled. To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step. This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature. There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice. In addition, researches often refer to the expected regret as \quotes{the regret}.}.
\eqn{
\label{eqn:raw_regret}
\rawregret = \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
}

@@

\vspace{0.5cm}
\begin{definition}[Pseudo-Regret]
\eqn{
\label{eqn:pseudo_regret}
\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}} \\
&= n\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{i}}}
}
\end{definition}

@@

\subsection{The counterfactual nature of regret}
\todo{Fix this bit up}

Note that we have used the conterfactual notation to represent the rewards given each action. The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep. The psuedo-regret (equation \ref{eqn:pseudo_regret}) remains unchanged as both it and the learner's actions depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$. However, the regret as defined in equation \ref{eqn:raw_regret} is no longer properly defined as it relies on both $\cfb{Y}{a_t}$ and $\cfb{Y}{i^*}$ being defined at each timestep. The variance (but not expectation) of $\rawregret$ depends on the covariance between counterfactual variables.

 The learner is only able to observe properties of the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
 The stochastic bandit problem can be defined without recourse to counterfactual variables by

 The algorithm and definition of regret depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.

@@

\subsubsection{The causal structure of contextual bandits}

\todo{think carefully about the graphs in this sections}
It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}. However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem. \citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information. This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
